---
title: "Child Birth Weight"
author: "Karim Reimoo"
date: "16/12/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source ("../R_src/child_birth_helper.R")
```


## 1 Pre-fitting data diagnostic
### 1.1 Original dataset summary
```{r}
chds <- read.csv("../data/chds_births.csv", sep=",")
summary(chds) 
```
$$ Table \ 1.1 $$

\newpage

### 1.2 Treatment of erroneous data
```{r}
# drop observations which have value 0 for covariate 'marital' 
chds <- chds[chds$marital != 0, ]
```

### 1.3 Treatment of NAs
```{r}
pairs(~ wt + fwt + fht, data=chds)
```
$$ Figure \ 1.1 $$

```{r, warning=FALSE}
library(mice)

# removing father's weight and height from our dataset 
chds$fwt <- NULL
chds$fht <- NULL

# imputing missing data
system.time({ 
  chds_imp = mice(chds, m=5, printFlag=FALSE, maxit=50, seed=331)
})

# extracting five imputed datasets
chds1 <- complete(chds_imp, 1)
chds2 <- complete(chds_imp, 2)
chds3 <- complete(chds_imp, 3)
chds4 <- complete(chds_imp, 4)
chds5 <- complete(chds_imp, 5)
```

\newpage

### 1.4 Converting categorical covariates into factors 
```{r}
# constructing smoke level factors
smoke_levels <- c("never", "until_pregnancy", "used_to_not_anymore", "smokes_now")

chds1 <- encode_data(chds1, smoke_levels, "smoke")
chds2 <- encode_data(chds2, smoke_levels, "smoke")
chds3 <- encode_data(chds3, smoke_levels, "smoke")
chds4 <- encode_data(chds4, smoke_levels, "smoke")
chds5 <- encode_data(chds5, smoke_levels, "smoke")

# constructing med factors
med_levels <- c("elementary", "middle", "high", "high+trade", "high+college", "college_grad",     
                "trade", "high_unclear")

chds1 <- encode_data(chds1, med_levels, "med")
chds2 <- encode_data(chds2, med_levels, "med")
chds3 <- encode_data(chds3, med_levels, "med")
chds4 <- encode_data(chds4, med_levels, "med")
chds5 <- encode_data(chds5, med_levels, "med")

# constructing fed factors
fed_levels <- c("elementary", "middle", "high", "high+trade", "high+college", "college_grad", 
                "trade", "high_unclear")


chds1 <- encode_data(chds1, fed_levels, "fed")
chds2 <- encode_data(chds2, fed_levels, "fed")
chds3 <- encode_data(chds3, fed_levels, "fed")
chds4 <- encode_data(chds4, fed_levels, "fed")
chds5 <- encode_data(chds5, fed_levels, "fed")

# constructiong marital factors
marital_levels <- c("married", "legally_separated", "divorced", "widowed", "never_married",                                   "not_married")
chds1 <- encode_marital(chds1, marital_levels, "martial", 0)
chds2 <- encode_marital(chds2, marital_levels, "martial", 0)
chds3 <- encode_marital(chds3, marital_levels, "martial", 0)
chds4 <- encode_marital(chds4, marital_levels, "martial", 0)
chds5 <- encode_marital(chds5, marital_levels, "martial", 0)

# constructing time factors 
time_levels <- c("never_smoked", "<1year", "1-2years", "2-3years", "3-4years", "5-9years", ">10years","during_pregnancy", "still_smokes", "quit_but_dont_know_when")

chds1 <- encode_data(chds1, time_levels, "time")
chds2 <- encode_data(chds2, time_levels, "time")
chds3 <- encode_data(chds3, time_levels, "time")
chds4 <- encode_data(chds4, time_levels, "time")
chds5 <- encode_data(chds5, time_levels, "time")

# constructing number factors 
number_levels <- c("never_smoked", "1-4", "5-9", "10-14", "15-19", "20-29", "30-39", "40-60", ">60","smoked_but_dont_know_how_much")

chds1 <- encode_data(chds1, number_levels, "number")
chds2 <- encode_data(chds2, number_levels, "number")
chds3 <- encode_data(chds3, number_levels, "number")
chds4 <- encode_data(chds4, number_levels, "number")
chds5 <- encode_data(chds5, number_levels, "number")

# constructing income factors
income_levels <- c("<2500", "2500-4999", "5000-7499", "7500-9999", "10000-12499", "12500-14999","15000-17499", "17500-19999", "20000-22499", ">=22500")

chds1 <- encode_data(chds1, income_levels, "income")
chds2 <- encode_data(chds2, income_levels, "income")
chds3 <- encode_data(chds3, income_levels, "income")
chds4 <- encode_data(chds4, income_levels, "income")
chds5 <- encode_data(chds5, income_levels, "income")

# constructing meth and feth factors
eth_levels <- c("Caucasian", "Mexican", "African-American", "Asian", "Mixed", "Other")

chds1 <- encode_eth_data(chds1)
chds2 <- encode_eth_data(chds2)
chds3 <- encode_eth_data(chds3)
chds4 <- encode_eth_data(chds4)
chds5 <- encode_eth_data(chds5)
```

\newpage

### 1.5 Handling covariate categories with few observations
```{r}
# Changing factor levels for marital
chds1 <- change_marital_factors(chds1)
chds2 <- change_marital_factors(chds2)
chds3 <- change_marital_factors(chds3)
chds4 <- change_marital_factors(chds4)
chds5 <- change_marital_factors(chds5)

# Changing factor levels for meth
chds1 <- change_meth_factor_levels(chds1)
chds2 <- change_meth_factor_levels(chds2)
chds3 <- change_meth_factor_levels(chds3)
chds4 <- change_meth_factor_levels(chds4)
chds5 <- change_meth_factor_levels(chds5)

# Changing factor levels for med
chds1 <- change_med_levels(chds1)
chds2 <- change_med_levels(chds2)
chds3 <- change_med_levels(chds3)
chds4 <- change_med_levels(chds4)
chds5 <- change_med_levels(chds5)

# Changing factor levels for feth
chds1 <- change_feth_levels(chds1)
chds2 <- change_feth_levels(chds2)
chds3 <- change_feth_levels(chds3)
chds4 <- change_feth_levels(chds4)
chds5 <- change_feth_levels(chds5)

# Changing factor levels for fed
chds1 <- change_fed_levels(chds1)
chds2 <- change_fed_levels(chds2)
chds3 <- change_fed_levels(chds3)
chds4 <- change_fed_levels(chds4)
chds5 <- change_fed_levels(chds5)

# Changing factor levels for income
chds1 <- change_income_levels(chds1)
chds2 <- change_income_levels(chds2)
chds3 <- change_income_levels(chds3)
chds4 <- change_income_levels(chds4)
chds5 <- change_income_levels(chds5)

# Changing factor levels for time
chds1 <- change_time_levels(chds1)
chds2 <- change_time_levels(chds2)
chds3 <- change_time_levels(chds3)
chds4 <- change_time_levels(chds4)
chds5 <- change_time_levels(chds5)


# Changing factor levels for time
chds1 <- change_number_levels(chds1)
chds2 <- change_number_levels(chds2)
chds3 <- change_number_levels(chds3)
chds4 <- change_number_levels(chds4)
chds5 <- change_number_levels(chds5)
```

\newpage

### 1.6 pairs plot for all the continuous covariates
```{r}
pairs(~ wt + gestation + parity + mage + mht + mwt + fage, data=chds1)
```

```{r}
library(plyr)
# counting distinct values in covariates "parity" and "mht"
count(chds1, vars="parity")
count(chds1, vars="mht")
```

\newpage
```{r}
# encode parity 
parity_levels <- c("0", "1", "2", ">=3")
parity_levels <- c("no_prev_preg", "one_prev_preg", "two_prev_preg", "at_least_three_prev_preg")

chds1 <- encode_parity(chds1)
chds2 <- encode_parity(chds2)
chds3 <- encode_parity(chds3)
chds4 <- encode_parity(chds4)
chds5 <- encode_parity(chds5)

# encode mht 
mht_levels <- c("at_most_60", "61-63", "64-66", "at_least_67")

chds1 <- encode_mht(chds1)
chds2 <- encode_mht(chds2)
chds3 <- encode_mht(chds3)
chds4 <- encode_mht(chds4)
chds5 <- encode_mht(chds5)
```

\newpage 
```{r}
# print comprehensive summary of factor encoded dataset
summary(chds1)
```
$$ Table \ 1.2 $$
\newpage

## 2 Model selection

### 2.1 Detecting and Removing Data Inconsistency
```{r}
# Removing inconsistencies between number and smoke
chds1 <- chds1[!(chds1$smoke == "smokes_now" & chds1$number == "never_smoked"), ]
chds1 <- chds1[!(chds1$smoke == "until_pregnancy" & chds1$number == "never_smoked"), ]

chds2 <- chds2[!(chds2$smoke == "until_pregnancy" & chds2$number == "never_smoked"), ]

chds4 <- chds4[!(chds4$smoke == "smokes_now" & chds4$number == "never_smoked"), ]
chds4 <- chds1[!(chds4$smoke == "until_pregnancy" & chds4$number == "never_smoked"), ]


table(chds1[c("number", "smoke")])
table(chds2[c("number", "smoke")])
table(chds3[c("number", "smoke")])
table(chds4[c("number", "smoke")])
table(chds5[c("number", "smoke")])
```

### 2.2 Automated model selection
```{r}

Mfwd1 <- forward_selection(chds1)
Mfwd2 <- forward_selection(chds2)
Mfwd3 <- forward_selection(chds3)
Mfwd4 <- forward_selection(chds4)
Mfwd5 <- forward_selection(chds5)

Mstep1 <- stepwise_selection(chds1)
Mstep2 <- stepwise_selection(chds2)
Mstep3 <- stepwise_selection(chds3)
Mstep4 <- stepwise_selection(chds4)
Mstep5 <- stepwise_selection(chds5)
```

### 2.3 Treatment of NAs in Beta_hat 
```{r}
# Extracting regression coeffecients
Beta_hat_Mfwd1 <- coef(Mfwd1)
Beta_hat_Mfwd2 <- coef(Mfwd2)
Beta_hat_Mfwd3 <- coef(Mfwd3)
Beta_hat_Mfwd4 <- coef(Mfwd4)
Beta_hat_Mfwd5 <- coef(Mfwd5)

Beta_hat_Mstep1 <- coef(Mstep1)
Beta_hat_Mstep2 <- coef(Mstep2)
Beta_hat_Mstep3 <- coef(Mstep3)
Beta_hat_Mstep4 <- coef(Mstep4)
Beta_hat_Mstep5 <- coef(Mstep5)

# Checking for any NAs in regression coeffcients
names(Beta_hat_Mfwd1)[is.na(Beta_hat_Mfwd1)]
names(Beta_hat_Mfwd2)[is.na(Beta_hat_Mfwd2)]
names(Beta_hat_Mfwd3)[is.na(Beta_hat_Mfwd3)]
names(Beta_hat_Mfwd4)[is.na(Beta_hat_Mfwd4)]
names(Beta_hat_Mfwd5)[is.na(Beta_hat_Mfwd5)]

names(Beta_hat_Mstep1)[is.na(Beta_hat_Mstep1)]
names(Beta_hat_Mstep2)[is.na(Beta_hat_Mstep2)]
names(Beta_hat_Mstep3)[is.na(Beta_hat_Mstep3)]
names(Beta_hat_Mstep4)[is.na(Beta_hat_Mstep4)]
names(Beta_hat_Mstep5)[is.na(Beta_hat_Mstep5)]
```

\newpage
### 2.4 Examining regression coeffecients in our fitted models
```{r}
formula(Mfwd1)
formula(Mfwd2)
formula(Mfwd3) 
formula(Mfwd4) 
formula(Mfwd5)

formula(Mstep1)
formula(Mstep2)
formula(Mstep3)
formula(Mstep4)
formula(Mstep5)
```

### 2.5 Selecting Best Forward and Stepwise Selection Models
```{r, warning=FALSE}
# We have four different models from forward and stepwise selection each 
# therefore, the following code is written  

# Stepwise Selection Final Model Competition
n_iter <- 2000
split <- 0.80

snames=expression(M[step1], M[step2], M[step3], M[step5])
fnames <- expression(M[fwd1], M[fwd2], M[fwd3], M[fwd5])
```

\newpage

```{r, warning=FALSE}
# Cross validation of unique step-wise selection models on chds1
cv(chds1, split, n_iter, Mstep1, Mstep2, Mstep3, Mstep5, snames, "chds1")
```
$$ Figure \ 2.5.1 $$

\newpage

```{r, warning=FALSE}
# Cross validation of unique step-wise selection models on chds2
cv(chds2, split, n_iter, Mstep1, Mstep2, Mstep3, Mstep5, snames, "chds2")
```
$$ Figure \ 2.5.2 $$

\newpage

```{r, warning=FALSE}
# Cross validation of unique step-wise selection models on chds3
cv(chds3, split, n_iter, Mstep1, Mstep2, Mstep3, Mstep5, snames, "chds3")
```
$$ Figure \ 2.5.3 $$
\newpage

```{r, warning=FALSE}
# Cross validation of unique step-wise selection models on chds4
cv(chds4, split, n_iter, Mstep1, Mstep2, Mstep3, Mstep5, snames, "chds4")
```
$$ Figure \ 2.5.4 $$
\newpage

```{r, warning=FALSE}
# Cross validation of unique step-wise selection models on chds5
cv(chds5, split, n_iter, Mstep1, Mstep2, Mstep3, Mstep5, snames, "chds5")
```
$$ Figure \ 2.5.5 $$
\newpage

```{r, warning=FALSE}
# Cross validation of unique forward selection models on chds1
cv(chds1, split, n_iter, Mfwd1, Mfwd2, Mfwd3, Mfwd5, fnames, "chds1")
```
$$ Figure \ 2.5.6 $$
\newpage

```{r, warning=FALSE}
# Cross validation of unique forward selection models on chds2
cv(chds2, split, n_iter, Mfwd1, Mfwd2, Mfwd3, Mfwd5, fnames, "chds2")
```
$$ Figure \ 2.5.7 $$
\newpage

```{r, warning=FALSE}
# Cross validation of unique forward selection models on chds3
cv(chds3, split, n_iter, Mfwd1, Mfwd2, Mfwd3, Mfwd5, fnames, "chds3")
```
$$ Figure \ 2.5.8 $$
\newpage

```{r, warning=FALSE}
# Cross validation of unique forward selection models on chds4
cv(chds4, split, n_iter, Mfwd1, Mfwd2, Mfwd3, Mfwd5, fnames, "chds4")
```
$$ Figure \ 2.5.9 $$
\newpage

```{r, warning=FALSE}
# Cross validation of unique forward selection models on chds5
cv(chds5, split, n_iter, Mfwd1, Mfwd2, Mfwd3, Mfwd5, fnames, "chds5")
```
$$ Figure \ 2.5.10 $$

### 2.6 Pooling Best Forward and Stepwise Selection Models
```{r}
# pooling Mfwd1
Mfwd_best <- with(chds_imp, 
                  lm(wt ~ gestation + smoke + mht + meth + parity + number + mwt + time + meth:mwt
                          + gestation:number + gestation:mwt + gestation:mht + gestation:time 
                    )
                  )
summary(pool(Mfwd_best))

# pooling Mstep2
Mstep_best <- with(chds_imp, 
                  lm(wt ~ gestation + parity + meth + med + mht + mwt + income + smoke + time +
                          number + gestation:income + gestation:number + gestation:med + med:mwt + meth:income
                    )
                  )
summary(pool(Mstep_best))
```

# 3 Model Diagnostics 

## 3.1 QQ & Residual Plots

```{r}

residual_vs_predicted_plot(Mfwd1, "Mfwd1")
qq_plot(Mfwd1, "Mfwd1")

residual_vs_predicted_plot(Mstep2, "Mstep2")
qq_plot(Mstep2, "Mstep2")
```


### 3.2 Leverage And Influence Analysis 
```{r}

cooks_distance_vs_leverage_plot(Mfwd1, chds1, "Mfwd1")
cooks_distance_vs_leverage_plot(Mstep2, chds2, "Mstep2")
```

## 3.3 Cook's High Influence vs High Leverage Plots
```{r, warning=FALSE}
#cooks_highinf_vs_highlev Mfwd1
p <- length(coef(Mfwd1))
n <- nrow(chds1)
h <- hatvalues(Mfwd1)
res <- resid(Mfwd1)
hbar <- p / n

# cook’s distance vs. leverage
D <- cooks.distance(Mfwd1)

# flag some of the points
infl_ind <- which.max(D) # top influence point
lev_ind <- h > 2*hbar # leverage more than 2x the average
clrs <- rep("black", len = n)

omit_ind <- c(infl_ind, # most influential
              which.max(h)) # highest leverage
names(omit_ind) <- c("infl", "lev")

yobs <- chds1[,"wt"] # observed values
 
# all observations
Mf <- lm(formula(Mfwd1), data=chds1)
yfitf <- predict(Mf) # fitted values
par(mfrow = c(2,2))
par(mar = c(3.5,3.5,0,0)+.1)
cex <- .8
clrs2 <- c("red", "blue")

for(jj in 1:length(omit_ind)) {
  Mo <- lm(formula(Mfwd1), data=chds1, subset = -omit_ind[jj])
  yfito <- predict(Mo, newdata = chds1) # fitted values at all observations
  for (ii in c("gestation", "mwt")) {
    # response vs. each covariate
    xobs <- chds1[,ii] # covariate
    ylim <- range(yobs, yfitf, yfito) # y-range of plot
    # observed
    plot(xobs, yobs, pch = 21, bg = clrs, cex = cex, cex.axis = cex, xlab = "", ylab = "")
    
    title(xlab = ii, ylab = "Weight", line = 2)
    # fitted, all observations
    points(xobs, yfitf, pch = 3, lwd = 2, cex = cex)
    # fitted, one omitted observation
    points(xobs, yfito, col = clrs2[jj], pch = 6, lwd = 2, cex = cex)
    # connect with lines
    segments(x0 = xobs, y0 = pmin(yobs, yfitf, yfito),
             y1 = pmax(yobs, yfitf, yfito), lty = 2) # connect them
    
    legend("bottomright", legend = c("Observed", "Fitted (all obs.)",
                                     paste0("Fitted (omit ", names(omit_ind)[jj], " obs.)")),
           pch = c(21,3,6), lty = c(NA, NA, NA),
           lwd = c(NA, 2, 2), pt.bg = "black",
           col = c("black", "black", clrs2[jj]),
           cex = cex, pt.cex = cex)
  }
}
```


```{r, warning=FALSE}
p <- length(coef(Mstep2))
n <- nrow(chds2)
h <- hatvalues(Mstep2)
res <- resid(Mstep2)
hbar <- p / n

# cook’s distance vs. leverage
D <- cooks.distance(Mstep2)

# flag some of the points
infl_ind <- which.max(D) # top influence point
lev_ind <- h > 2*hbar # leverage more than 2x the average
clrs <- rep("black", len = n)

omit_ind <- c(infl_ind, # most influential
              which.max(h)) # highest leverage
names(omit_ind) <- c("infl", "lev")

yobs <- chds2[,"wt"] # observed values
Mf <- lm(formula(Mstep2), data=chds2)
yfitf <- predict(Mf) # fitted values
par(mfrow = c(2,2))
par(mar = c(3.5,3.5,0,0)+.1)
cex <- .8
clrs2 <- c("red", "blue")

for(jj in 1:length(omit_ind)) {
  # model with omitted observation
  Mo <- lm(formula(Mstep2), data=chds2, subset = -omit_ind[jj])
  yfito <- predict(Mo, newdata = chds2) # fitted values at all observations
  for (ii in c("gestation", "mwt")) {
    # response vs. each covariate
    xobs <- chds2[,ii] # covariate
    ylim <- range(yobs, yfitf, yfito) # y-range of plot
    # observed
    plot(xobs, yobs, pch = 21, bg = clrs, cex = cex, cex.axis = cex, xlab = "", ylab = "")
    
    title(xlab = ii, ylab = "Weight", line = 2)
    # fitted, all observations
    points(xobs, yfitf, pch = 3, lwd = 2, cex = cex)
    # fitted, one omitted observation
    points(xobs, yfito, col = clrs2[jj], pch = 6, lwd = 2, cex = cex)
    # connect with lines
    segments(x0 = xobs, y0 = pmin(yobs, yfitf, yfito),
             y1 = pmax(yobs, yfitf, yfito), lty = 2) # connect them
    
    legend("bottomright", legend = c("Observed", "Fitted (all obs.)",
                                     paste0("Fitted (omit ", names(omit_ind)[jj], " obs.)")),
           pch = c(21,3,6), lty = c(NA, NA, NA),
           lwd = c(NA, 2, 2), pt.bg = "black",
           col = c("black", "black", clrs2[jj]),
           cex = cex, pt.cex = cex)
  }
}
```
